Stability Analysis of Recurrent Neural Networks \\ using Ordinary Differential Equations

### 2020-07-12 p5/stability_rnn.tex
First draft of attention layer.

\documentclass{article}
\usepackage{amsmath, amssymb, amsfonts, amsthm}
\numberwithin{equation}{section}
\newtheorem{thm}{Theorem}
\newtheorem{defn}{Definition}
\newtheorem*{rem}{Remark}
\begin{document}
\title{Stability Analysis of Recurrent Neural Networks \\
    using Ordinary Differential Equations}
\author{Author's name}
\maketitle
\section{Introduction}\label{s1}
Recurrent neural networks feed their output to their input. This feature
makes them look like algorithms to get numerical solutions of ordinary
differential equations. It is well known that training neural networks is
rife with vanishing and exploding gradients\cite{gereon2018hands}. This
phenomenon is closely related to the behaviour of the numerical algorithms
to solve ordinary differential equations, a topic that has been studied
intensively for several decades. We use this analogy to bank upon the 
vast research about ordinary differential equations to derive criteria 
for the stability of recurrent neural networks.

@@

\noindent We also define
\begin{defn}\label{s7d2}
The Burrage-Butcher tensor corresponding to the $(n,t)$-ODERNN of 
definition \ref{s7d1} is
\begin{equation}\label{x}
Q_{i, j} = \beta_i\alpha_{i, j} + \beta_j\alpha_{j, i} - \beta_i\beta_j^T.
\end{equation}
\end{defn}

@@

\section{GRU with an Attention Layer}\label{s10}
We will now analyze GRU with attention layer. We assume that the 
architecture of the network is same as shown in figure 4 of the paper
\cite{ran2019lstm} by Xiangdong and others, except that the LTSM is 
replaced with a GRU. There are $n$ such GRU's, each with an output
$h^i_l \in \mathbb{R}^s$. The combined output of all of them is the $s
\times n$ matrix $Y_l$ defined by
\begin{equation}\label{s10e1}
H_l = [h^1_l, \ldots, h^n_l].
\end{equation}
The matrix $H_l$ is scaled with an $n \times 1$  weight matrix $W_H$. 
Similarly, if $A$ is an $s \times n$ embedding matrix and $W_v$ is the
corresponding weight matrix, then
\begin{equation}\label{s10e2}
M_l = \tanh\begin{pmatrix} H_lW_H \\ AW_v \end{pmatrix}
\end{equation}
is a member of $\mathbb{R}^{2s}$. If $W= [w_1, \ldots, w_n]$ is a $(2s) 
\times n$ weight matrix then
\begin{equation}\label{s10e3}
a_i = \frac{\exp(w_i^T M_l)}{\sum_{j=1}^n \exp(w_j^T M_l)}
\end{equation}
are the $n$ components of the vector $a = (a_1, \ldots, a_n)
\in \mathbb{R}^n$. Finally, the product
\begin{equation}\label{s10e4}
Y_l = H_l a
\end{equation}
is a member of $\mathbb{R}^s$. Note that the output of the individual
GRUs is also an $s$-dimensional vector.

@@

Equation \eqref{s10e4} tells that $Y_l$ depends $H_l$ but not the $H_{l-1}$.
Thus, it is not a recurrence relation. $Y_l$ is not computed from the
`present' $H_l$, not a `previous' one. Therefore, this attention layer 
does not play a role in determining the stability of the network in the
architecture shown in figure 4 of the paper \cite{ran2019lstm} by 
Xiangdong and others.

>>>>

\documentclass{article}
\usepackage{amsmath, amssymb, amsfonts, amsthm}
\DeclareMathOperator*{\argmax}{argmax}
\DeclareMathOperator*{\smax}{softmax}
\numberwithin{equation}{section}
\newtheorem{thm}{Theorem}
\newtheorem{defn}{Definition}
\newtheorem*{rem}{Remark}
\begin{document}
\title{Stability Analysis of Recurrent Neural Networks \\
    using Ordinary Differential Equations}
\author{Author's name}
\maketitle
\section{Introduction}\label{s1}
Recurrent neural networks feed their output to their input. This feature
makes them look like algorithms to get numerical solutions of ordinary
differential equations. It is well known that training neural networks is
rife with vanishing and exploding gradients\cite{gereon2018hands}. This
phenomenon is closely related to the behaviour of the numerical algorithms
to solve ordinary differential equations, a topic that has been studied
intensively for several decades. We use this analogy to bank upon the 
vast research about ordinary differential equations to derive criteria 
for the stability of recurrent neural networks.

@@

\noindent We also define
\begin{defn}\label{s7d2}
The Burrage-Butcher tensor corresponding to the $(n,t)$-ODERNN of 
definition \ref{s7d1} is
\begin{equation}\label{s7e4}
Q_{i, j} = \beta_i\alpha_{i, j} + \beta_j\alpha_{j, i} - \beta_i\beta_j^T.
\end{equation}
\end{defn}

@@

\section{GRU with an Attention Layer}\label{s10}
We will now analyze GRU with attention layer used in three separate
architectures. 
\subsection{Architecture proposed Ran and others} 
We assume that the architecture of the network is same as shown in figure 4 
of the paper \cite{ran2019lstm} by Ran and others, except that the 
Long Short-Term Model (LTSM) is replaced with a GRU. There are $n$ such 
GRU's, each with an output $h^i_l \in \mathbb{R}^s$. The combined output 
of all of them is the $s
\times n$ matrix $Y_l$ defined by
\begin{equation}\label{s10e1}
H_l = [h^1_l, \ldots, h^n_l].
\end{equation}
The matrix $H_l$ is scaled with an $n \times 1$  weight matrix $W_H$. 
Similarly, if $A$ is an $s \times n$ embedding matrix and $W_v$ is the
corresponding weight matrix, then
\begin{equation}\label{s10e2}
M_l = \tanh\begin{pmatrix} H_lW_H \\ AW_v \end{pmatrix}
\end{equation}
is a member of $\mathbb{R}^{2s}$. If $W= [w_1, \ldots, w_n]$ is a $(2s) 
\times n$ weight matrix then
\begin{equation}\label{s10e3}
a_i = \frac{\exp(w_i^T M_l)}{\sum_{j=1}^n \exp(w_j^T M_l)}
\end{equation}
are the $n$ components of the vector $a = (a_1, \ldots, a_n)
\in \mathbb{R}^n$. Finally, the product
\begin{equation}\label{s10e4}
Y_l = H_l a
\end{equation}
is a member of $\mathbb{R}^s$. Note that the output of the individual
GRUs is also an $s$-dimensional vector.

@@

Equation \eqref{s10e4} tells that $Y_l$ depends $H_l$ but not the $H_{l-1}$.
Thus, it is not a recurrence relation. $Y_l$ is not computed from the
`present' $H_l$, not a `previous' one. Therefore, this attention layer 
does not play a role in determining the stability of the network in the
architecture shown in figure 4 of the paper \cite{ran2019lstm} by 
Ran and others.

\subsection{Architecture proposed by Zang and others}
We now consider the model proposed by Zang and others\cite{zhang2017gru}.
The model is described by equations (11) to (17) in their paper. We
reproduce them for an easy reference.
\begin{eqnarray}
z^\prime_t &=& \sigma(W_{yz}Ey_{t-1}+U_{sz}s_{t-1}+C_{cz}c_t)\label{s10e5}\\
r^\prime_t &=& \sigma(W_{yr}Ey_{t-1}+U_{sr}s_{t-1}+C_{cr}c_t)\label{s10e6}\\
\tilde{s}_t &=& \tanh\left(W_{ys}Ey_{t-1} + U_{rs}(r^\prime_t \circ
	    s_{t-1} + C_{cs}c_t\right) \label{s10e7} \\
s_t &=& (1 - z^\prime_t)\circ s_{t-1} + z^\prime_t \circ \tilde{s}_t
    \label{s10e8} \\
e_{ti} &=& \nu^T_{\text{att}}\tanh(W_{\text{att}}s_{t-1} +
    U_\text{att}a_i \label{s10e9} \\
\alpha_{ti} &=& 
    \frac{\exp(e_{ti})}{\sum_{k=1}^L\exp(e_{tk})}\label{s10e10}\\
c_t &=& \sum_{i=1}^L \alpha_{ti}a_i. \label{s10e11}
\end{eqnarray}
In these equations the lower case symbols are either vectors or scalars;
the upper case symbols are matrices. In particular:
\begin{itemize}
\item $y_i \in \mathbb{R}^K$;
\item $a_i \in \mathbb{R}^D$;
\item $s_i, \tilde{s}_i, z_i, r_t \in \mathbb{R}^n$;
\item $W_{yz}, W_{yr}, W_{ys}, U_{sz}, U_{sr}, U_{rs}, C_{cz}, C_{cr}, 
C_{cs}$ are weight matrices of appropriate dimensions. $E$ is an embedding
matrix;
\item If $n^\prime$ is the attention dimension then $\nu_{\text{att}}^{n
^\prime}$, $W_\text{att}$ is an $n \times n^\prime$ matrix and $U_a$ is an 
$n^\prime \times D$ matrix.
\end{itemize}
The attention layer determines the vector $c_t$, which is analogous to
the bias vectors $b_q$, $q = 1, \ldots, n + 1$ in equations \eqref{s7e1}
to \eqref{s7e3} in definition \ref{s7d1}. The bias vectors \emph{do not}
contribute to the Burrage-Butcher tensor defined in \eqref{s7e4}. Therefore,
we conclude that the attention layer has no influence on the stability of
the underlying network.

\subsection{Architecture proposed by Liu and others}
In this section we consider the neural network used in table-to-text
generation by Liu and others \cite{liu2018table}. The goal of their
network is to generate a sequence
\begin{equation}\label{s10e12}
w^\ast_{1:p} = \argmax_{w_{1:p}}\prod_{t=1}^p\mathrm{P}(w_t | w_{0:t-1}, 
	R_{1:n}),
\end{equation}
where $w_i$ are generated tokens and $R_n$ are field value records in a
table. The probability on the right hand side is defined by
\begin{equation}\label{s10e13}
\mathrm{P}(w_t| w_{0:t-1}, R_{1:n}) = \smax(W_s \circ g_t),
\end{equation}
where $W_s$ is a weight matrix and
\begin{equation}\label{s10e14}
g_t = \tanh(W_t[s_t, a_t])
\end{equation}
$s_t$ being the output of an LTSM neural network and $a_t$ being the 
attention vector. It is built as
\begin{equation}\label{s10e15}
a_t = \sum_{i=1}^L \alpha_{t_i}h_i,
\end{equation}
where
\begin{equation}\label{s10e16}
\alpha_{t_i} = \frac{\exp(g(s_t, h_i))}{\sum_{j=1}^N\exp(g(s_t, h_j))},
\end{equation}
and
\begin{equation}\label{s10e17}
g(s_t, h_i) = \tanh(W_p h_i) \circ \tanh(W_q s_t).
\end{equation}
Here $s_t$ is the output of the LTSM at time $t - 1$ and 
\begin{equation}\label{s10e18}
h_i = o_t \circ \tanh(c_t).
\end{equation}
The variables in this equation are boolean vectors of size $n$. They 
are defined as
\begin{equation}\label{s10e19}
\begin{pmatrix}i_t \\ f_t \\ o_t \\ \hat{c}_t \end{pmatrix} = 
\begin{pmatrix}\sigma \\ \sigma \\ \sigma \\ \tanh\end{pmatrix}
W_{4n, 2n}^c\begin{pmatrix}d_t \\ h_{t-1} \end{pmatrix}
\end{equation}
and
\begin{equation}\label{s10e20}
c_t = f_t \circ c_{t-1} + i_t \circ \hat{c}_t.
\end{equation}
Analogous to the case of the architecture proposed by \cite{ran2019lstm}
the attention layer $a_t$ is used solely that the final stage. It is not
fed back to the input and therefore it has no influence on the stability
of the LTSM.

\subsection{Remarks on stability}
Recall theorem \ref{s9t1} that the stability of the Runge-Kutta method 
depends on the whether or not the matrices $\beta_n$ \emph{and} the 
Burrage-Butcher tensor are positive semi-definite. An attention layer will
influence the stability of an algorithm if it influences either $\beta_i$
of $\alpha_{i,j}$ which together form the matrices that determine the
stability. The architectures of Ran et al. and Liu et al. use the 
attention layer only at the last stage. They do not feed the attention
later at time $t$ to get the output at time $t + 1$. On the other hand, 
the architecture of Zang et al. uses the attention layer in bias vectors
which do not determine the matrices that decide the stability of the
algorithm.

The stability of Runge-Kutta methods depends on their ability to generate
contractive solutions in an interval. If a method is not stable it does not
mean that it cannot generate \emph{any} contractive pair of solutions. All
it means is that \emph{not all} pairs are contractive in the sense of 
definition \ref{s9d2}. Existence of one, or a few, solutions to the 
algorithm is not a contradiction to it being unstable. Instability means
that \emph{not all} solutions are stable, and stability means that
\emph{all} solutions are stable. An existence of a few coverging solutions
does not tell us anything about the stability of the algorithm. However, 
an existence of a single unstable solution renders an algorithm unstable.
